---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sigstats

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/sigstats)](https://CRAN.R-project.org/package=sigstats)
[![R-CMD-check](https://github.com/selkamand/sigstats/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/selkamand/sigstats/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**sigstats** enables common mathematical operations / transformations to be applied to **sigverse** style signatures / catalogues

## Installation

You can install the development version of sigstats like so:

``` r
if (!require("pak", quietly = TRUE))
    install.packages("pak")

pak::pak("selkamand/sigstats")
```

## Quick Start


```{r fig.width=10, fig.height=4}
library(sigstats)
library(sigstash)
library(sigvis)

# Load a signature collection
signatures <- sig_load("COSMIC_v3.3.1_SBS_GRCh38")

# Create a model that represents a mix of SBS2 (40%) and SBS13 (60%)
model <- c(SBS2 = 0.4, SBS13 = 0.6)

# Add selected signatures to the combined model
combined_signatures <- sig_combine(signatures, model)

# Visualise result
sig_visualise(
  combined_signatures, 
  title = "Model",
  subtitle = "Created by combining SBS2 (40%) and SBS13 (60%)"
  )
```

### Signature Operations

sigstats helps you add and subtract catalogues/signatures

```{r}
# Load a signature collection
signatures <- sig_load("COSMIC_v3.3.1_SBS_GRCh38")

# Reconstruct catalogues for two pure samples (each with 100 mutations)
catalogue1 <- sig_reconstruct(signatures[['SBS3']], n = 100)
catalogue2 <- sig_reconstruct(signatures[['SBS4']], n = 100)
catalogue3 <- sig_reconstruct(signatures[['SBS5']], n = 100)

# Subtract catalogue2 from catalogue1
difference <- catalogue1 %-% catalogue2

# Inspect result
head(difference)

# Sum three catalogues
catalogue1 %+% catalogue2 %+% catalogue3

# Sum a catalogue collection
catalogues <- list(cat1 = catalogue1, cat2 = catalogue2, cat3 = catalogue3)
sig_sum(catalogues)
```



### Reconstruct a mutation catalogue from a signature model

We often need to reconstruct a catalogue (or tally) from our signature model.

```{r fig.width=10, fig.height=4}
# Load a signature collection
signatures <- sig_load("COSMIC_v3.3.1_SBS_GRCh38")

# Create a model that represents a mix of SBS2 (40%) and SBS13 (60%)
model <- c(SBS2 = 0.4, SBS13 = 0.6)

# Create a new signature by combining SBS2 and SBS13 in ratios dictated by the above model
signature <- sig_combine(signatures, model)

# Reconstruct a perfect catalogue describing what the mutational profile of a sample
# with 200 mutations and the given signature model would look like
reconstuction <- sig_reconstruct(signature, n=200)

# Visualise result
sig_visualise(
  reconstuction, 
  class = "catalogue",
  title = "Reconstructed Catalogue",
  subtitle = "Expected profile of a sample with 200 mutations: 40% from SBS2, 60% from SBS13"
  )
```


### Compute Stats on Signature Collections

```{r}

# Load a signature collection
signatures <- sig_load("COSMIC_v3.3.1_SBS_GRCh38")

# Compute common statistics:
# e.g. Gini coefficient, Exponentiated Shannon Index, KL divergence, L1/L2/L3 Norms)
stats <- sig_collection_stats(signatures)

# Print Stats
head(stats)

# Plot kl_divergence_from_uniform against exponentiated shannon index
plot(
  x = stats[["kl_divergence_from_uniform"]], 
  y = stats[["shannon_index_exp_scaled"]]
)
```
### Compute Similarity/Distance Measures

```{r}
signatures <- sig_load("COSMIC_v3.3.1_SBS_GRCh38")

# Pairwise cosine similarity
sim_df <- sig_collection_pairwise_stats(
  signatures,
  metric = "cosine_similarity",
  format = "data.frame"
)
head(sim_df)

# L2 (Euclidean) distance matrix
dist_mat <- sig_collection_pairwise_stats(
  signatures,
  metric = "L2",
  format = "matrix"
)
```


### Model Correctness

Quantify how well a fitted model matches ground truth signature weights.

```{r}

signatures <- sig_load("COSMIC_v3.3.1_SBS_GRCh38")
cosmic_signatures <- names(signatures)

observed <- c(SBS1 = 0.7, SBS5 = 0.3, SBS18 = 0)
truth <- c(SBS1 = 0.6, SBS5 = 0.4, SBS18 = 0)

metrics <- sig_model_correctness(observed, truth, all_signatures = cosmic_signatures)

print(metrics)
```

